---
title: "Moodys Calculation"
output: html_notebook
---

# Introduction
We will apply a Granger causality test to see if we can replicate Moodys' results published in their February 17 Newsletter.

# Research Problem
Here is the story from [Moodys, page 5](#r1).

>Income sentiment could mislead

>Persistently strong inflation requires consumers to expect
higher incomes that allow them to finance these price
increases. A potential silver lining for the Fed is that the net
percentage of consumers expecting higher incomes in the
next six months has fallen. Historically, this measure of
income sentiment is strongly correlated with the Atlanta
Fed's Wage Growth Tracker. We used a Granger causality
test to determine if there is a causal relationship between
the two. With various lags, income sentiment was found to
Granger-cause changes in the Atlanta Fed wage tracker. The
causal relationship runs in only one direction.

We could not access the Consumer Confidence Index, prepared by The Conference Board, without paying $995.00. So we went for the [University of Michigan's Index of Consumer Sentiment (ICS)](#r2)   The Atlanta Fed Wage Tracker (AFWT) is publicly available, and we obtained it.

# Packages
We load the standard MBAD packages.
```{r}
library(tidyverse) # used to manipulate data and plot
library(ggfortify) # some extra ggplot functions
library(GGally) # pairs ggplot and other options
library(skimr) # nice summaries
library(broom) # neat model summary
library(gt) # nice displays of printed data
library(patchwork) # arranging graphs
library(car) # useful tools for regression analysis
library(yardstick) # for model performance metrics
library(ggResidpanel) # for diagnostic residual plots
```

# Custom Packages for MBAD
Install and load:
```{r}
# allow package installation from github
library(devtools) # Make sure that the devtools library is loaded
if(!("mbadtools" %in% installed.packages())) devtools::install_github("bhartman2/mbadtools", force=T)
# load github packages
library(mbadtools)
```

# Data
The next sections show how to get the data for the Moody report. But we have already installed the data required inthe `mbadtools` package as dataset `moody`.

<!---
## Atlanta Fed Wage Tracker
First we load this dataset.
```{r}
library(readxl)
wage_growth_data <- read_excel("../inst/wage-growth-data.xlsx", 
    skip = 1)
glimpse(wage_growth_data)
```

## UM Consumer Confidence
```{r}
library(readxl)
redbk01a <- read_excel("../inst/redbk01a.xls")
glimpse(redbk01a)
```

# Data Description
Fed Wage Growth Data:
```{r}
wage_growth_data
```

```{r}
redbk01a
```

# Data Wrangling

## AFWT
```{r warning=FALSE}
waget = wage_growth_data %>% select(1,2) %>% rename(Date=1, Wgrowth=2) %>%
  mutate(Wgrowth=as.numeric(Wgrowth)) %>% na.omit %>%
  mutate(Month = lubridate::month(Date), Year=lubridate::year(Date))
glimpse(waget)
```

Visualizing:
```{r}
waget %>% ggplot() +
  geom_point(aes(Date, Wgrowth)) +
  geom_line(aes(Date, Wgrowth))
```


## ICS
```{r}
umics = redbk01a %>% 
  rename(Month=`Date  of  Survey`, Year=2) %>% na.omit %>% 
  mutate(Mo = match(Month, month.name)) %>%
  mutate(Date=paste0(Year,"-",Mo,"-01"), 
         Datep=as.POSIXct(as.Date(Date))%m+%months(1))
glimpse(umics)
umics
```

```{r}
umics %>% ggplot() + 
  geom_point(aes(Datep,ICS)) +
  geom_line(aes(Datep,ICS))
```

And together:
```{r}
ggplot(umics %>% dplyr::filter(Datep>=waget[1,]$Date)) + 
  geom_point(aes(Datep,ICS, color="Consumer Confidence")) +
  geom_line(aes(Datep,ICS), linetype="dotted") +
  geom_point(aes(Date,Wgrowth*28, color="Wage Growth"), data=waget) +
  geom_line(aes(Date,Wgrowth*28), data=waget, linetype="dotted") +
  # labs(y="ICS and Wage Growth*25") +
  scale_color_manual(values=c("red","black"), name="")+
  theme(legend.position = "bottom") +
  scale_y_continuous(name = "Consumer Confidence Index", 
    sec.axis = sec_axis(~./28, name = "Wage Growth*25")) 
```
It seems like Wage growth starts latest, so we must select only data in its data range for inclusion.
```{r}
waget[1,]$Date
```

```{r}
umics1 = umics %>% 
  dplyr::filter(Datep>=waget[1,]$Date) %>% 
  select(Datep, ICS) #%>% 
umics2 = umics1 %>% mutate(Datep=Datep + days(1) + hours(6) + minutes(59))
waget1 = waget %>% select(Date, Wgrowth)
data.frame(umics2, waget1)
```
--->

```{r}
# moody = data.frame(waget, umics1) %>% select(Date, Wgrowth, ICS)
data(moody, package="mbadtools")
glimpse(moody)
```

```{r}
ggplot(moody) + 
  geom_point(aes(Date,ICS, color="Consumer Confidence")) +
  geom_point(aes(Date,Wgrowth*25, color="Wage Growth")) +
  # labs(y="ICS and Wage Growth*25") +
  scale_color_manual(values=c("red","black"), name="")+
  theme(legend.position = "bottom") +
  scale_y_continuous(name = "Consumer Confidence Index", 
    sec.axis = sec_axis(~./25, name = "Wage Growth*25")) 
```
It seems that consumer confidence drops hard just after wage growth peaks.

# Granger Causality
We will test Granger causality between these two series, using 4 lags.

## `ICS` =G=> `Wgrowth`
```{r}
lmtest::grangertest(Wgrowth~ICS, data=moody, order=4)
```

Yes, it appears `ICS` does Granger-cause `Wgrowth`.

## `Wgrowth` =G=> `ICS`
```{r}
lmtest::grangertest(ICS~Wgrowth, data=moody, order=4)
```

`Wgrowth` does not Granger-cause `ICS.`

# `mbadtools` functions
We will tune the order parameter by choosing different lags of 1 through 10 for the Granger-causality test. We'll run both directions of causality, and plot the results. Package `mbadtools` contains functions to test and plot for all lags up to a number given by `max.lags`, which defaults to 10.

```{r}
# test Granger Causality
gpv = moody %>% GrangerTestPvals(ICS, Wgrowth, max.lags=10)
gpv
```

We can plot this dataset with another function from `mbadtools`.
```{r fig.height=9, fig.width=12}
gpv %>% GrangerPlot() %>% wrap_plots(nrow=2)
```

Clearly ICS =G=> Wgrowth, with better than .05 p-value for lags 1, 4, 7, and 8.  And for all lags, p-value is below .10.

On the other hand, Wgrowth !=G=> ICS for any tested lag, with all p-values exceeding .25.

# EU Stock Markets
This dataset from the `datasets` package contains time series of four indices from the EU stockmarkets. Does any of them Granger-cause another? We will test `FTSE` and `DAX`.

```{r}
data(EuStockMarkets)
# Convert to data frame for easy handling
df <- as.data.frame(EuStockMarkets)
df
```

Perform Granger causality test (e.g., DAX causing FTSE and the reverse)

```{r}
gt = GrangerTestPvals(df, FTSE, DAX)
gt
```

Plot the results
```{r fig.height=10, fig.width=10}
gt %>% GrangerPlot() %>% wrap_plots(nrow=2)
```

# Conclusion
Our result confirms what [the Moodys article](#r1) says. Consumer confidence Granger-causes Wage growth, but not the reverse. This is telling, because consumer confidence should lead to more spending, and that would generate wage growth. 

We also tested FTSE and DAX indices from the `EUStockMarkets` dataset. We found that FTSE Granger-causes DAX, but not the reverse.

# References
::: {#r1}
1. https://www.moodys.com/researchdocumentcontentpage.aspx?docid=PBC_1319860
:::

:::{#r2}
2. https://data.sca.isr.umich.edu/tables.php
:::