---
title: "Pareto Plots"
author:
  name: "Bruce Hartman"
  affiliation: "University of St. Francis"
date: "`r lubridate::today()`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    theme: united
abstract: |
  We demonstrate Pareto Plots from the `qcc` package and then test our own functions for them based on `ggplot`.
---

# Pareto Plots
Pareto plots are a good way to understand what's important in data with categories. 

# Packages
```{r}
library(tidyverse) # used to manipulate data and plot
library(ggfortify) # some extra ggplot functions
library(GGally) # pairs ggplot and other options
library(skimr) # nice summaries
library(broom) # neat model summary
library(gt) # nice displays of printed data
library(patchwork) # arranging graphs
library(car) # useful tools for regression analysis
library(yardstick) # for model performance metrics
library(ggResidpanel) # for diagnostic residual plots
```

# Custom Packages for MBAD
Install and load:
```{r}
# allow package installation from github
library(devtools) # Make sure that the devtools library is loaded
if(!("mbadtools" %in% installed.packages())) devtools::install_github("bhartman2/mbadtools")
# load github packages
library(mbadtools)
```

# Package `qcc`
```{r}
library(qcc)
```

# Data 
Package `qcc` has several datasets with examples of Pareto Plots.
```{r}
data(mtcars) # from qcc package
mtcars
cyl_count <- table(mtcars$cyl)
pareto.chart(cyl_count, main = "Pareto Chart of Car Cylinders")
```

`HairEycolor` is stored as a table.
```{r}
data(HairEyeColor)
HairEyeColor %>% str()
HairEyeColor
eye_count <- apply(HairEyeColor, c(2), sum) #2 refers to eyes
pareto.chart(eye_count, main = "Pareto Chart of Eye Colors")
```

`Titanic` is stored as a table.
```{r}
data(Titanic)
Titanic
class_count <- apply(Titanic, c(1), sum) #1 refers to Class
pareto.chart(class_count, main = "Pareto Chart of Passenger Classes")
```

`defects` is stored as an ordered vector of counts.
```{r}
defects <- c("Scratches" = 40, "Dents" = 25, "Misalignment" = 15, "Discoloration" = 10, "Other" = 5)
pareto.chart(defects, main = "Pareto Chart of Product Defects")
```

These plots are not structured with `ggplot`. But they work.

# The `mbadtools` Method
We use the data management and plotting functions in my package. All begin with `pareto_`.

First step is to create a summarized dataset.
```{r}
A = mtcars %>% mutate(cyl= as.character(cyl))
A1 = A %>% group_by(cyl) %>% summarize(c_count=n())
A1
```
Can apply `pareto_arrange_data()` function to A1.

```{r}
A2 = A1 %>% pareto_arrange_data(c_count)
A2
```
Noiw the basic plot.
```{r}
A2 %>% pareto_plot(cyl, c_count)
```

Add second axis label:
```{r}
mc = max(A2$c_count)
A2 %>% pareto_plot(cyl, c_count) +
  pareto_sec_axis(mc)
```
Add percent labels on the cumulative percent line. You can control them with `nudge_y` and `size`.
```{r}
mc = max(A2$c_count)
A2 %>% pareto_plot(cyl, c_count) +
  pareto_sec_axis(mc) +
  pareto_pct_labels(mc, nudge_y=1, size=3)
```

Add a percent cutoff line and plot labeling:
```{r}
mc = max(A2$c_count)
A2 %>% pareto_plot(cyl, c_count) +
    pareto_sec_axis(mc) +
    pareto_pct_labels(mc, nudge_y=.8, size=3) +
    pareto_cutoff_line(mc) +
    labs(title = "Pareto Chart of cylinder counts",
       x = "No of Cylinders",
       y = "Frequency")
```

This is what we want our Pareto chart to look like, with all the attachments. You may not want both the right hand axis and the cutoff line.

# Examples from ChatGPT
I had ChatGPT find some data examples that would work for Pareto plots.

## US Cities
```{r}
data("us.cities", package="maps")  # U.S. cities dataset
# us.cities
U = us.cities %>% group_by(country.etc) %>% 
  summarize(cpop = sum(pop))
U1 = U %>% pareto_arrange_data(cpop)
U1
```

```{r}
U1 %>% head(19) %>% 
  pareto_plot(country.etc, cpop) + 
  labs(title = "Pareto Chart of Top U.S. States by Population",
       x = "State",
       y = "Population Count") +
  pareto_pct_labels(max(U1$cpop), size=2.7) +
  pareto_sec_axis(max(U1$cpop)) +
  pareto_cutoff_line(max(U1$cpop)) +
  geom_text(aes(x=country.etc, y=cpop/2,
                label = round(cpop/1e6,1))
            ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```
19 states make up 80% of the US population.

## NYC Flights
```{r}
library(nycflights13)
# Group by delay reason and count occurrences
delay_data <- flights %>% 
  dplyr::filter(dep_delay > 0) %>%
  mutate(delay_category = case_when(
    dep_delay < 15 ~ "Short Delay",
    dep_delay < 60 ~ "Medium Delay",
    TRUE ~ "Long Delay"
  )) %>%
  group_by(delay_category) %>%
  summarise(count = n()) 
delay_data

D1 = delay_data %>% pareto_arrange_data(count)
mx=max(D1$count)
D1 %>% pareto_plot(delay_category, count) +
  pareto_pct_labels(mx, nudge_y=1000) +
  pareto_sec_axis(mx) +
  geom_text(aes(x=delay_category,
                y=count/1.5,
                label = paste0("~",round(count,-2)))
            ) +
  labs(title = "Pareto Chart of Flight Delays",
       x = "Delay Category") +
  theme_minimal()

```

## Logistics
Container flows for countries from 2010 thru 2019.

```{r}
library(readxl)
ContainerFlows <- read_excel(
  "../inst/API_IS.SHP.GOOD.TU_DS2_en_excel_v2_1657.xls",
  skip = 3)
ContainerFlows %>% skim()
```

Select the years with most nations' data complete.
```{r}
CF = ContainerFlows %>% select(`Country Name`, `Country Code`, 
                               `2010`, `2011`, `2012`, `2013`, `2014`,
                               `2015`, `2016`, `2017`, `2018`, `2019`)
CF %>% skim
```

Pivot longer.
```{r}
CF1 = CF %>% pivot_longer(cols=3:12, names_to = "Year", 
                          values_to = "Flow")
CF1 %>% glimpse
```

Make Year nueric and rename `Country Code` and `Country Name` columns.
```{r}
CF2 = CF1 %>% mutate(Year = as.numeric(Year)) %>%
  rename(Country=`Country Code`, CName=`Country Name`)
CF2 %>% glimpse
```

```{r}
CF2 %>% skim
```

Remove missing data.
```{r}
CF3 = CF2 %>% na.omit
CF3 %>% skim
```

Make Pareto data frame: use mean of flow over the 10 years for each country. Change mean to millions of containers.
```{r}
CF4 = CF3 %>% group_by(Country) %>% 
  summarize(cflow = mean(Flow, na.rm=T)/1e6) 
CF4 %>% glimpse
```

Filter only the Country column items that correspond to countries, rather than other statistical units. Display in a `gt` table.
```{r}
CF4
df = dplyr::filter(CF4,
  !(Country %in% c("CEB","CSS","EAP","EAR",
                 "EAS","ECS","EMU","EUU",
                 "FCS","HIC","IBD","IBT",
                 "IDA","IDB","LAC","LCN",
                 "LMC","LMY","LTE","MEA",
                 "MIC","MNA","NAC","OED",
                 "OSS","PRE","PSS","PST",
                 "SSA","SST","TEA","TEC",
                 "TLA","TMN","TSA","TSS",
                 "UMC","WLD","ARB","AFW",
                 "SSF")) 
              )

df %>% gt() %>% fmt_number(decimals=3) %>% opt_interactive()
```

Get data ready and make Pareto Plot.
```{r}
df1 = df %>% pareto_arrange_data(cflow)
df1
mc = max(df1$cflow)
```

Pick a number of countries for about 80% and make Pareto Plot.
```{r}
q=25
C = df1 %>% head(q) %>% 
  pareto_plot(Country, cflow) +
  labs(title=paste0("Pareto Plot of Flow, ",q," largest Countries"),
       x="Country Code",
       y="Average Trade Flow 2010-2019") +
  theme(axis.text.x = element_text(angle=45, hjust=1)) +
  # pareto_pct_labels(mc) +
  pareto_cutoff_line(mc) +
  pareto_sec_axis(mc)
C
```

We don't know country codes, so extract the names and make a dataframe with `Country` and `CName`.
```{r}
a = df$Country %>% enframe(value="Country")

b = data.frame(cn = CF3 %>% distinct(CName), 
               Country = CF3 %>% distinct(Country)
               )
c = left_join(a, b, by="Country")
w = c %>% select(-name)
ww = w %>% dplyr::filter(Country %in% (df1 %>% head(q) %>% pull(Country)) )
```

Add the table to the plot display. Note that we do not have to make the table of countries a `gt` table; `wrap_table` does that for us. We may have to change the height and width of the plot, and we specify a design to make the Pareto plot 75% of the plot width.

```{r fig.height=9, fig.width=12}
list(C, W=wrap_table(ww)) %>% wrap_plots(design = "CCCW") 
```

# Example for package
```{r}
data(chickwts, package="datasets")
K = chickwts %>% group_by(feed) %>% summarize(avwt = mean(weight))
K1 = pareto_arrange_data(K, avwt)
K1 %>% pareto_plot(feed, avwt) +
  labs(x="feed",
       y="average weight")
```

# Conclusion
Tests show the Pareto plots are working.
