---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `mbadtools`

<!-- badges: start -->
<!-- badges: end -->

The goal of `mbadtools` is to provide MBAD students with the key packages the course focuses on, and to also provide some tools (in the form of functions) for certain tasks often encountered in your work. 

## Installation

You can install the current version of `mbadtools` like so:

```{r}
if(!("mbadtools" %in% installed.packages())) 
  pak::pak("bhartman2/mbadtools")
```

## Notebook or Script Usage

This is a basic example which shows you how to load `mbadtools`; that is, make `mbadtools` available to your notebook. The initial output of this example is shown here, but you cn suppress it in your notebook; it's only informative if you have a problem.

```{r example, echo=TRUE}
library(mbadtools)
```

Because `lmtest` overwrites functions in package `dplyr` (a part of tidyverse), we reset the most important conflicts to prioritize `dplyr` over others.

```{r}
# Tell R to always prefer dplyr's select or filter function over any other package
conflicted::conflict_prefer_matching("select|filter", "dplyr", quiet=T)
```

## Packages Loaded
Here is a current list of the packages loaded with `mbadtools`.

| Package | Use |
|:-------- | ---------- |
| tidyverse | host of packages for modern data and code handling in R |
| ggfortify | tools to improve use of `ggplot2` graphics |
| GGally | tools for `ggplot2` graphics especially the `ggpairs()` function |
| skimr | better than `summary()` for quickly exploring and summarizing data |
| gt | displaying neat, readable tabular data |
| patchwork | tools for combining `ggplot` objects |
| car | tools for statistical modeling and data analysis |
| lmtest | tools for advanced statistical modeling and data analysis |
| yardstick | tools for model comparison and metrics |
| gh4x | for the `geom_pointpath()` function |
| ggpubr | for the `ggqqplot()` function |


## Basic Notebook Template

A basic notebook template is provided to help you set up a notebook to write an assignment. It is an .Rmd file like a miniature paper with code and output interspersed, and with typical key headings you may want to include.

[View Basic Notebook Template](https://htmlpreview.github.io/?https://github.com/bhartman2/mbadtools/blob/main/notebook/Basic%20Notebook%20Template.nb.html)

## Functions in `mbadtools`

Some useful functions are included in `mbadtools`.

| Function | Description  |
|:-------- | :----------- |
|`gg_residual_plots()` | displays 7 different residual plots for linear regression 
|`gg_partial_residual_plots()` | displays partial residual plots for linear regression | " |
|`gt_add_significance()` | creates a `gt` table for a dataframe with a `p.value` column; adds color when p.value is significant |
|||
|`GrangerTest()` | performs Granger causality tests for time series data in a data frame |
|`GrangerPlot()`| plots results of `GrangerTest()` |
|`GrangerTestPvals()` | performs Granger causality tests for time series data |
|`GrangerTestTune()`| plots results of `GrangerTestPvals()` |

The first 3 are demonstrated in [Using mbadtools](https://htmlpreview.github.io/?https://github.com/bhartman2/mbadtools/blob/main/notebook/Using%20mbadtools.nb.html).

The Granger Causality functions are demonstrated in [Granger MoodysDemo](https://htmlpreview.github.io/?https://github.com/bhartman2/mbadtools/blob/main/notebook/GrangerMoodysDemo.nb.html)


Others can be viewed in the `mbadtools` help.



