---
title: "Moodys Calculation"
output: html_notebook
---

# Introduction
We will apply a Granger causality test to see if we can replicate Moodys' results published in their February 17 2026 Newsletter.

# Research Problem
Here is the story from [Moodys, page 5](#r1).

>Income sentiment could mislead

>Persistently strong inflation requires consumers to expect
higher incomes that allow them to finance these price
increases. A potential silver lining for the Fed is that the net
percentage of consumers expecting higher incomes in the
next six months has fallen. Historically, this measure of
income sentiment is strongly correlated with the Atlanta
Fed's Wage Growth Tracker. We used a Granger causality
test to determine if there is a causal relationship between
the two. With various lags, income sentiment was found to
Granger-cause changes in the Atlanta Fed wage tracker. The
causal relationship runs in only one direction.

We could not access the Consumer Confidence Index, prepared by The Conference Board, without paying $995.00. So we went for the [University of Michigan's Index of Consumer Sentiment (ICS)](#r2)   The Atlanta Fed Wage Tracker (AFWT) is publicly available, and we obtained it.

# Packages
We load the standard MBAD packages.

# Custom Packages for MBAD
Install and load:
```{r}
library(mbadtools)
lib_packages(mbad_standard_packages)
library(lmtest)
```

# Data
The next sections show how to get the data for the Moody report. But we have already installed the data required inthe `mbadtools` package as dataset `moody`.

## Atlanta Fed Wage Tracker
First we load this dataset.
```{r}
library(readxl)
wage_growth_data <- read_excel("../inst/wage-growth-data.xlsx", 
    skip = 1)
glimpse(wage_growth_data)
```

## UM Consumer Confidence
```{r}
library(readxl)
redbk01a <- read_excel("../inst/redbk01a.xls")
glimpse(redbk01a)
```

# Data Description
Fed Wage Growth Data:
```{r}
wage_growth_data
```

```{r}
redbk01a
```

# Data Wrangling

## AFWT
```{r warning=FALSE}
waget = wage_growth_data %>% select(1,2) %>% rename(Date=1, Wgrowth=2) %>%
  mutate(Wgrowth=as.numeric(Wgrowth)) %>% na.omit %>%
  mutate(Month = lubridate::month(Date), Year=lubridate::year(Date))
glimpse(waget)
```

Visualizing:
```{r}
waget %>% ggplot() +
  geom_pointpath(aes(Date, Wgrowth)) 
```


## ICS
```{r}
umics = redbk01a %>% 
  rename(Month=`Date  of  Survey`, Year=2) %>% na.omit %>% 
  mutate(Mo = match(Month, month.name)) %>%
  mutate(Date=paste0(Year,"-",Mo,"-01"), 
         Datep=as.POSIXct(as.Date(Date))%m+%months(1))
glimpse(umics)
umics
```

```{r}
umics %>% ggplot() + 
  geom_pointpath(aes(Datep,ICS))
```

And plotting together:
```{r}
ggplot(umics %>% dplyr::filter(Datep>=waget[1,]$Date)) + 
  geom_pointpath(aes(Datep,ICS, color="Consumer Confidence")) +
  geom_pointpath(aes(Date,Wgrowth*28, color="Wage Growth"), 
                 data=waget) +
  # labs(y="ICS and Wage Growth*25") +
  scale_color_manual(values=c("red","black"), name="")+
  theme(legend.position = "bottom") +
  scale_y_continuous(name = "Consumer Confidence Index", 
    sec.axis = sec_axis(~./25, name = "Wage Growth*25")) 
```
It seems like Wage growth starts latest, so we must select only data in its data range for inclusion.
```{r}
waget[1,]$Date
```

```{r}
umics1 = umics %>% 
  dplyr::filter(Datep>=waget[1,]$Date) %>% 
  select(Datep, ICS) #%>% 
umics2 = umics1 %>% mutate(Datep=Datep + days(1) + hours(6) + minutes(59))
waget1 = waget %>% select(Date, Wgrowth)
data.frame(umics2, waget1)
```

# The `moody` Data
```{r}
data(moody, package="mbadtools")
glimpse(moody)
```

```{r}
ggplot(moody) + 
  geom_pointpath(aes(Date,ICS, color="Consumer Confidence")) +
  geom_pointpath(aes(Date,Wgrowth*25, color="Wage Growth")) +
  # labs(y="ICS and Wage Growth*25") +
  scale_color_manual(values=c("red","black"), name="")+
  theme(legend.position = "bottom") +
  scale_y_continuous(name = "Consumer Confidence Index", 
    sec.axis = sec_axis(~./25, name = "Wage Growth*25")) 
```
It seems that consumer confidence drops hard just after wage growth peaks. The `moody` data matches the data gleaned form the other sources. 

# Granger Causality
We will test Granger causality between these two series, using 4 lags, with the `lmtest::grangertest()` function. Its limitation is that you have to do one lag at a time, and one direction at a time.

## `ICS` =G=> `Wgrowth`
```{r}
lmtest::grangertest(Wgrowth~ICS, data=moody, order=4)
```

Yes, it appears `ICS` does Granger-cause `Wgrowth`.

## `Wgrowth` =G=> `ICS`
```{r}
lmtest::grangertest(ICS~Wgrowth, data=moody, order=4)
```

`Wgrowth` does not Granger-cause `ICS.`

# Granger Causality with `mbadtools`
We will tune the order parameter, which represents the number of lags, by choosing different lags of 1 through 10 (or a user-chosen number) for the Granger-causality test. We'll run both directions of causality, and plot the results. Package `mbadtools` contains functions to test and plot for all lags up to a number given by `max.lags`, which defaults to 10.
 
Test the `moody` data with `wgrowth` and `ICS`.
```{r}
moody %>% str
```

The dataset is a data frame. so we use the data frame functions from `mbadtools`.
```{r fig.height=7}
moody %>% GrangerTable(ICS, Wgrowth) %>% 
  GrangerPlot() %>% patchwork::wrap_plots(nrow=2)
```

We can also do it as a time series.
```{r fig.height=7}
ICSt = moody %>% select(ICS) %>% ts
Wgrt = moody %>% select(Wgrowth) %>% ts
GrangerTestPvals(ICSt, Wgrt)
GrangerTestTune(ICSt, Wgrt) %>% patchwork::wrap_plots(nrow=2)
```

Clearly ICS =G=> Wgrowth, with better than .05 p-value for lags 1, 4, 7, and 8.  And for all lags, p-value is below .10.

On the other hand, Wgrowth !=G=> ICS for any tested lag, with all p-values exceeding .25.

# EU Stock Markets
This dataset from the `datasets` package contains time series of four indices from the EU stockmarkets. Does any of them Granger-cause another? We will test `FTSE` and `DAX`.

Perform Granger causality test (e.g., `DAX` causing `FTSE` and the reverse), using a data frame: 

```{r fig.height=7}
data(EuStockMarkets)
EuStockMarkets %>% glimpse
# Convert to data frame for non-time-series handling.
EuStockMarkets %>% data.frame %>% 
  GrangerTable(DAX, FTSE) %>% 
  GrangerPlot() %>% patchwork::wrap_plots(nrow=2)
```

Using time-series directly: columns of `EuStockMarkets` are time series.
```{r fig.height=7}
DAX = EuStockMarkets[,"DAX"]
FTSE = EuStockMarkets[,"FTSE"]
GrangerTestPvals(DAX, FTSE)
GrangerTestTune(DAX, FTSE) %>% 
  wrap_plots(nrow=2)
```

# Conclusion
Our result confirms what [the Moodys article](#r1) says. Consumer confidence Granger-causes Wage growth, but not the reverse. This is telling, because consumer confidence should lead to more spending, and that would generate wage growth. 

We also tested FTSE and DAX indices from the `EUStockMarkets` dataset. We found that FTSE Granger-causes DAX, but not the reverse.

# References
::: {#r1}
1. https://www.moodys.com/researchdocumentcontentpage.aspx?docid=PBC_1319860
:::

:::{#r2}
2. https://data.sca.isr.umich.edu/tables.php
:::